Assessing spatial patterning of Middle Paleolithic human settlement in Western Iberia

### 2020-10-15 analysis/paper/paper.Rmd
rmarkdown start

---
title: "Title Goes Here"
author:
  - Author One
  - Author Two
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

library(MidPalSet) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

@@

Here is a citation [@Marwick2017]

# Background

```{r demo, eval = FALSE}
1 + 1
```

@@

@@

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

>>>>

---
title: "Assessing spatial patterning of Middle Paleolithic human settlement in Western Iberia"
author:
  - João Cascalheira
  - Célia Gonçalves
  - Daniela Maio
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  Middle Paleolithic; GIS; Western Iberia
---

@@

#library(MidPalSet) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory

library(rgdal)
library(raster)
library(rasterVis)
library(lattice)
library(foreach)
library(rnaturalearth)
library(tidyverse)
library(ggplot2)
library(broom)
library(foreach)
library(sf)

```

```{r load-data, eval = FALSE, echo = FALSE}

#Download geo tif from OSF repositorium

#osf_retrieve_file("nv6r8") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("jb6v8") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("yxk23") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("57qfn") %>%
#  osf_download(path = "./analysis/data/raw_data")
#osf_retrieve_file("mx2bt") %>%
#  osf_download(path = "./analysis/data/raw_data")

#Import geo tiffs
areaDEM <- raster("../data/raw_data/areaDEM.tif")
areaDEM_cut <- raster("../data/raw_data/areaDEM_cut.tif")
areaSLOPE_cut <- raster("../data/raw_data/areaSLOPE_cut.tif")
areaASPECT_cut <- raster("../data/raw_data/areaASPECT_cut.tif")
areaDIST_cut <- raster("../data/raw_data/areaDIST_cut2.tif")

# Convert coordinates to longlat WGS84
areaDEM <- projectRaster(areaDEM, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaDEM_cut <- projectRaster(areaDEM_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaSLOPE_cut <- projectRaster(areaSLOPE_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaASPECT_cut <- projectRaster(areaASPECT_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
areaDIST_cut <- projectRaster(areaDIST_cut, crs="+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

#Import sites
sites <- read.csv("../data/raw_data/sites_short_list.csv", header=TRUE)
caves <- read.csv("../data/raw_data/all_caves.csv", header=TRUE)

#Convert to shapefile
coordinates(sites)<-~Longitude+Latitude
proj4string(sites) = CRS("+init=epsg:4326")
writeOGR(sites, "../data/raw_data/spatial_data/shapefiles","sites_short", "ESRI Shapefile")

coordinates(caves)<-~Longitude+Latitude
proj4string(caves) = CRS("+init=epsg:4326")
writeOGR(caves, "../data/raw_data/spatial_data/shapefiles","caves", "ESRI Shapefile")

#Read shapefiles
sites_short <- readOGR(dsn="../data/raw_data/spatial_data/shapefiles", layer="sites_short")
caves <- readOGR(dsn="../data/raw_data/spatial_data/shapefiles", layer="caves")
#rivers <- readOGR(dsn="../data/raw_data/spatial_data/shapefiles", layer="RHidro_Rivers_v2")

#Filter sites by Type
sites_sub <- sites_short[sites_short$Type == "Cave" | sites_short$Type == "Open air",]
sites_sub$Type <- factor(sites_sub$Type)

#Convert coordinates to longlat WGS84
sites_sub_long <- spTransform(sites_sub, "+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

caves_long <- spTransform(caves, "+proj=longlat +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

```

```{r extract-variables, eval = FALSE, echo = FALSE}

# Extract variables

# Resample Distance raster
areaDIST_cut <- resample(areaDIST_cut,areaDEM_cut, resample='bilinear')

# Stack maps
terrainstack <- stack(areaDEM_cut,
                      areaSLOPE_cut,
                      areaASPECT_cut,
                      areaDIST_cut)

# Extract values from stack for each site within a 20 m radius
sites_vals <- raster::extract(terrainstack,
                      sites_sub_long,
                      buffer = 20,
                      fun = mean,
                      sp = TRUE)

# Subdivide sites for further analysis
open_air <- dplyr::filter(as.data.frame(sites_vals), Type == "Open air")
cave <- dplyr::filter(as.data.frame(sites_vals), Type == "Cave")

```

@@

```{r general-map, eval = FALSE, echo = FALSE}

rasterVis::levelplot(areaDEM,
                     margin = list(x = FALSE,
                                   y = TRUE),
                     col.regions = terrain.colors(16),
                     xlab = list(label = "",
                                 vjust = -0.25),
                     sub = list(
                       label = "Meters a.s.l.",
                       font = 1,
                       cex = .9,
                       hjust = 0.5),
                     colorkey=list(space="bottom"),
                     key = list(
                       space = "left",
                       points = list(
                         pch = c(18,20),
                         col = c("red","blue")),
                       text = list(
                         c("Cave","Open-air"),
                         cex=1))) +
  spplot(sites_sub_long, # add a layer of points
         zcol = "Type",
         cex = 2,
         pch = c(18,20),
         col.regions = c("red","blue"))

```

@@

## Site selection

## Mapping and GIS analysis

## Statistical analysis

@@

```{r plot-altitude, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_altitude_densities <- open_air %$%
   areaDEM_cut %>%
   density(from = -26, to = 1993, n = 1965) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 1965) %>%
   dplyr::rename(Elevation = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_altitude_values <- areaDEM_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_altitude_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_altitude_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = -26, to = 1993, n = 1965) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 1965)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Elevation", "Lower CI", "Frequency", "Upper CI"))

# Plot distributions
ggplot() +
  geom_line(data = background_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency)) +
  geom_ribbon(data = background_altitude_densities,
              mapping = aes(x = Elevation,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_altitude_densities,
            mapping = aes(x = Elevation,
                          y = Frequency),
            color = "red")

```

```{r test-altitude, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_altitude_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_altitude_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaDEM_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)

# Get the median test statistic and 95% confidence interval
background_altitude_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_altitude_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))

round(background_altitude_MWU, 3)

```

```{r plot-slope, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_slope_densities <- open_air %$%
   areaSLOPE_cut %>%
   density(from = 0, to = 70, n = 35) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 35) %>%
   dplyr::rename(Slope = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_slope_values <- areaSLOPE_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_slope_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_slope_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = 0, to = 70, n = 35) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 35)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Slope", "Lower CI", "Frequency", "Upper CI"))

# Plot distributions
ggplot() +
  geom_line(data = background_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency)) +
  geom_ribbon(data = background_slope_densities,
              mapping = aes(x = Slope,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_slope_densities,
            mapping = aes(x = Slope,
                          y = Frequency),
            color = "red")

```

```{r test-slope, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_slope_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_slope_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaSLOPE_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)

# Get the median test statistic and 95% confidence interval
background_slope_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_slope_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))

round(background_slope_MWU, 3)

```

```{r plot-aspect, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_aspect_densities <- open_air %$%
   areaASPECT_cut %>%
   density(from = -1, to = 360, n = 181) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 181) %>%
   dplyr::rename(Aspect = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_aspect_values <- areaASPECT_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_aspect_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_aspect_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = -1, to = 360, n = 181) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 181)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Aspect", "Lower CI", "Frequency", "Upper CI"))

# Plot distributions
ggplot() +
  geom_line(data = background_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency)) +
  geom_ribbon(data = background_aspect_densities,
              mapping = aes(x = Aspect,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_aspect_densities,
            mapping = aes(x = Aspect,
                          y = Frequency),
            color = "red")

```

```{r test-aspect, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_aspect_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_aspect_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaASPECT_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)

# Get the median test statistic and 95% confidence interval
background_aspect_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_aspect_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))

round(background_aspect_MWU, 3)

```

```{r plot-distance, eval = FALSE, echo = FALSE}

# Calculate site altitude densities
open_air_distance_densities <- open_air %$%
   areaDIST_cut %>%
   density(from = 0, to = 7697, n = 350) %>%
   broom::tidy() %>%
   tibble::as_tibble() %>%
   dplyr::mutate(y = y * 350) %>%
   dplyr::rename(Distance = x,
                Frequency = y)

# Load background values into memory for fast resampling
background_distance_values <- areaDIST_cut %>%
  values() %>%
  na.omit()

# Draw 1000 random samples from background and calculate their densities
background_distance_densities <- foreach::foreach(n = 1:1000, .combine = rbind) %do% {
  background_distance_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    density(from = 0, to = 7697, n = 350) %>%
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 350)
} %>%
  dplyr::group_by(x) %>%
  purrrlyr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "cols") %>%
  magrittr::set_names(c("Distance", "Lower CI", "Frequency", "Upper CI"))

# Plot distributions
ggplot() +
  geom_line(data = background_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency)) +
  geom_ribbon(data = background_distance_densities,
              mapping = aes(x = Distance,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = open_air_distance_densities,
            mapping = aes(x = Distance,
                          y = Frequency),
            color = "red")

```

```{r test-aspect, echo = false, eval = FALSE}

# Draw 1000 random samples from background, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
background_distance_MWU <- foreach(n = 1:1000, .combine = rbind) %do% {
  background_sample <- background_distance_values %>%
    sample(nrow(open_air),
           replace = FALSE) %>%
    wilcox.test(x = open_air$areaDIST_cut,
                y = .,
                exact = FALSE) %>%
    broom::tidy() %>%
    tibble::as_tibble()

} %>%
  dplyr::select(statistic, p.value)

# Get the median test statistic and 95% confidence interval
background_distance_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  background_distance_MWU %>%
    dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))

round(background_distance_MWU, 3)

```
